Containers and R---the Rockerverse and beyond

### 2020-01-16 manuscript.Rmd
add section about Gigantum image stack, add @davclark, extend conclusion
closes #29

Co-authored-by: Dav Clark <dav@gigantum.com>

@@

**[Gigantum](http://gigantum.com/)** also its own image stack ([https://github.com/gigantum/base-images](https://github.com/gigantum/base-images)).
...

@@

The R packages and use cases presented show the positive drive in the community to innovate, but also the challenge of keeping up with a continuously evolving landscape.
The use cases contributed by co-authors also have a degree of overlap, which can be expected as a common language and understanding of good practices is still taking shape.
Also, the ease with which one can create a complex software systems, such as an independent Docker image stack, to serve one's specific needs leads to some fragmentation.
Furthermore fragmentation may not be a bad sign but instead a reflection of a growing market, which is able to sustain multiple related efforts.
With the maturing of core building blocks, such as the Rocker suite of images, more systems will be built successfully but will also be behind the curtains.
Docker alone, as a flexible core technology, is not a feasible level of collaboration and abstraction.
Instead, the use cases and applications observed in this work provide a more useful division.

Still, at least _on the level of R packages_ some consolidation seems in order, e.g., to reduce the number of packages creating `Dockerfile`s from R code or controlling the Docker daemon with R code.
It remains to be seen which approach to control Docker, via the Docker API as `stevedore` or via system calls as `dockyard`/`docker`/`dockr`, is more sustainable, or if the question will be answered by the endurance of maintainers and sufficient funding.
Similarly, the capturing of environments and their serialization in form of a `Dockerfile` currently happens at different levels of abstraction and re-use of functionality seems reasonable, e.g. `liftr` could generate the environment with `containerit`, which in turn may use `dockerfiler` for low level objects representing a `Dockerfile`.
In this consolidation, the Rocker&nbsp;Project could play the role of a unifying and coordinating entity in the future.
Though for the moment, the sign of the times points to more experimentation and feature growth, e.g. images for GPU-based computing and artificial intelligence.
Even with coding being more and more accepted as a required, and achievable skill, an easier access, for example by exposing containerisation benefits via simple user interfaces in the users' IDE, could be an important next step.
Currently containerisation happens more in the background at the system level.

@@

DN conceived of the presented idea and \href{https://github.com/nuest/rockerverse-paper/issues/3}{initialised the formation of the writing team}, wrote sections not mentioned below, and revised all sections.
DE wrote the introduction and section about Rocker.
CB ..
RC and EH wrote the section on interfaces for Docker in R.
ME ..
CF wrote the paragraphs about `r-online`, `dockerfiler`, `r-ci` and `r-db`.
BW wrote the section on research compendia and made the project Binder-ready.
KR ..
NR ..
NX wrote the section on \pkg{liftr}.
LS \& NT wrote the section on Bioconductor.
DB wrote the section on \pkg{outsider}.
All authors approved the final version.

>>>>

  - name: Dav Clark
    # https://github.com/davclark
    affiliation: Gigantum, Inc.
      - 1140 3rd Street NE
      - Washington, D.C. 20002, USA
      - \orcid{0000-0002-3982-4416}
    email: dav@gigantum.com

@@

**Gigantum** ([http://gigantum.com/](http://gigantum.com/) is a platform for open and decentralized data science with a focus on using automation and user-friendly tools for easy sharing of reproducible computational workflows.
Gigantum builds on the locally running _Gigantum CLI_ and _Gigantum Desktop_ applications for development and execution of analyses, which can be stored and shared via _Gigantum Cloud_.
The Gigantum apps are user friendly interfaces to a backend using Docker containers to package a Gigantum project ([https://github.com/gigantum/base-images](https://github.com/gigantum/base-images)).
The available images include two with R based on Ubuntu Linux have the [`c2d4u` CRAN PPA](https://launchpad.net/~marutter/+archive/ubuntu/c2d4u3.5/) pre-configured for installation of binary R packages\footnote{\href{https://docs.gigantum.com/docs/using-r}{https://docs.gigantum.com/docs/using-r}}.
<!-- mention lack of littler? -->
The R images vary in the included authoring environment, i.e., Jupyter in `r-tidyverse` or both Jupyter \& RStudio in `rstudio-server`.
The independent image stack can be traced back to the Gigantum environment and its features.
The R images are based on Gigantum's `python3-minimal` image, originally to keep the existing front-end configuration, but also to provide a consistent Python-to-R interoperability.
The `Dockerfile`s also use build args to specify bases, for example for different version of NVIDIA CUDA for GPU processing\footnote{See \href{https://github.com/gigantum/base-images/blob/master/_templates/python3-minimal-template/Dockerfile}{https://github.com/gigantum/base-images/blob/master/_templates/python3-minimal-template/Dockerfile} for the `Dockerfile` of `python3-minimal`.} so that appropriate GPU drivers can be enabled automatically when supported.
Furthermore, Gigantum's focus lies on environment management via GUI and ensuring a smooth user interaction, e.g., with reliable and easy conflict detection and resulution.
<!-- @DC: when a user interactively installs packages in Python, do you write that to the Dockerfile? UI-based environment management does not work for R packages, right? \href{https://docs.gigantum.com/docs/environment-management}{https://docs.gigantum.com/docs/environment-management} -->
Because of these specifics, instructions from `rocker/r-ubuntu` could _not_ be readily re-used in this image stack (see Section&nbsp;\nameref{conclusion}).
The image build and publication process is scripted with Python and JSON template configuration files, unlike Rocker which relies on more plain `Dockerfile`s.
A minor reason in the inception of the images were also project constraints requiring a Rocker-incompatible licensing of the `Dockerfile`s, i.e., the MIT License.

@@

The R packages and use cases presented show the positive drive in the community to innovate, but also the challenge of keeping up with a continuously evolving landscape.
The use cases contributed by co-authors also have a degree of overlap, which can be expected as a common language and understanding of good practices is still taking shape.
Also, the ease with which one can create a complex software systems with Docker, such as an independent Docker image stack, to serve one's specific needs leads to parallel developments.
<!-- Tyler W./DC from Gigantum: ideas about ease-of-DIY together with difficulty of composition; shared starting points of Rocker -->
This ease-of-DIY in combination with the difficulty to compose environments based on `Dockerfile`s alone is a further reason for **fragmentation**.
Instructions can be outsourced into distributable scripts and then copied into the image during build, but that make `Dockerfile` hard to read and adds a layor of complexity.
Despite the different image stacks presented here, the pervasiveness of Rocker can be traced back to its maintainers and the user community valueing collaboration and shared starting points over impulses to create individual solutions.
Aside from that, fragmentation may not be a bad sign, but instead a reflection of a growing market, which is able to sustain multiple related efforts.
With the maturing of core building blocks, such as the Rocker suite of images, more systems will be built successfully but will also be behind the curtains.
Docker alone, as a flexible core technology, is not a feasible level of collaboration and abstraction.
Instead, the use cases and applications observed in this work provide a more useful division.

Nonetheless, at least on the level of R packages some **consolidation** seems in order, e.g., to reduce the number of packages creating `Dockerfile`s from R code or controlling the Docker daemon with R code.
It remains to be seen which approach to control Docker, via the Docker API as `stevedore` or via system calls as `dockyard`/`docker`/`dockr`, is more sustainable, or if the question will be answered by the endurance of maintainers and sufficient funding.
Similarly, the capturing of environments and their serialization in form of a `Dockerfile` currently happens at different levels of abstraction and re-use of functionality seems reasonable, e.g. `liftr` could generate the environment with `containerit`, which in turn may use `dockerfiler` for low level objects representing a `Dockerfile`.
In this consolidation, the Rocker&nbsp;Project could play the role of coordinating entity.
Though for the moment, the sign of the times points to more experimentation and feature growth, e.g. images for GPU-based computing and artificial intelligence.
Even with coding being more and more accepted as a required, and achievable skill, an easier access, for example by exposing containerisation benefits via simple user interfaces in the users' IDE, could be an important next step.
Currently containerisation happens more in the background at the system level.

@@

DN conceived of the presented idea and \href{https://github.com/nuest/rockerverse-paper/issues/3}{initialised the formation of the writing team}, wrote sections not mentioned below, and revised all sections.
DE wrote the introduction and section about Rocker.
CB ..
RC and EH wrote the section on interfaces for Docker in R.
ME ..
CF wrote the paragraphs about `r-online`, `dockerfiler`, `r-ci` and `r-db`.
BW wrote the section on research compendia and made the project Binder-ready.
KR ..
NR ..
NX wrote the section on \pkg{liftr}.
LS \& NT wrote the section on Bioconductor.
DB wrote the section on \pkg{outsider}.
DC contributed content on Gigantum.
All authors approved the final version.
